---
title: "Introduction to Lee-Carter"


execute:
  echo: false
  eval: true
  collapse: false
  message: false
  warning: false
  code-fold: false


engine: knitr
format:
  pdf:
    output-file: lab-lee-carter.pdf
  html:
    output-file: lab-lee-carter.html


---

{{< include _preamble.qmd >}}



```{r}
#| label: setup
#| warning: false
#| message: false
#| echo: false
#| appendix: false

require(arrow)
require(demography)
require(tidyverse)
require(gt)
require(skimr)
require(GGally)
require(broom)
require(patchwork)
require(ggforce)
require(glue)
require(ggfortify)
require(ggforce)

tidymodels::tidymodels_prefer(quiet = TRUE)

old_theme <-theme_set(theme_bw(base_size=9, base_family = "Helvetica"))
```

[tidyverse conflicts](https://tidyverse.tidyverse.org/reference/tidyverse_conflicts.html)

```{r}
#| echo: false
#| appendix: false
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  comment=NA,
  prompt=FALSE,
  cache=FALSE,
  echo=TRUE,
  results='asis'
)
```




```{r}
#| appendix: false
params <- list(
  country_code= 'FRA',
  country= 'France',
  datafile= '~/Dropbox/HMD/hld-part.parquet',
  year_p= 1948,
  year_e= 2020
)
```

## Data


Life data tables are downloaded from $https://www.mortality.org](https://www.mortality.org).
and [https://www.lifetable.de](https://www.lifetable.de), see [Lab on life tables](/core/labs/lab-lt-1948-2016.qmd). If you install and load package [https://cran.r-project.org/web/packages/demography/index.html](https://cran.r-project.org/web/packages/demography/index.html), you will also find life data tables.
We investigate life tables describing countries from Western Europe (France, Great Britain, Italy, the Netherlands, Spain, Portugal, and Sweden), Canada, and the United States.

We load the one-year lifetables for females and  males for the different countries.


```{r}
#| label: country-codes
#| appendix: true

country_iso3 <- c(
  "France"="FRA",
  "Canada"="CAN",
  "Netherlands"="NLD",
  "Sweden"="SWE",
  "Italy"="ITA",
  "Spain"="ESP",
  "Portugal"="PRT",
  "USA"="USA"
)

tb_iso3 <- tibble(
  name=names(country_iso3), iso3=country_iso3
)
```

```{r}
#| label: col-selection
#| appendix: true

ds_path <- '~/Dropbox/HMD/hld-part.parquet'

ds <- open_dataset(ds_path)
```

```{r}
#| label: load-lt-countries
#| appendix: true
tb <- ds |>
  filter(
    Country %in% country_iso3, 
    Sex %in% c(1L, 2L), 
    Region=='0', 
    Residence=='0',
    SocDem=='0',
    Version=='1',
    Year1 >= 1948, 
    Ethnicity=='0', 
    AgeInt==1,
    TypeLT==1) |> 
    select(-c(Region, Ethnicity,Year2,AgeInt, TypeLT, Residence, SocDem, Version)) |> 
  collect()
```

```{r}
#| label: crude-averaging
#| appendix: true
tb <- tb  |>  
  summarise(across(ends_with('(x)'), mean), .by=c(Country, Sex, Year1, Age)) 
```


```{r}
#| appendix: true
life_table <- tb |>
  mutate(Iso3 = factor(Country)) |>
  mutate(Gender = factor(Sex, labels=c("Male", "Female"))) |>
  mutate(Area = fct_collapse(Iso3, 
    SE = c("PRT", "ESP", "ITA", "FRA"), 
    NE = c("NLD", "SWE", "CAN"), 
    USA = "USA")) |> 
  rename(Year=Year1) |> 
  select(Area, Country, Gender, Year, Age, `q(x)`) |> 
  mutate(`q(x)`=ifelse(`q(x)`==0, 2e-5 , `q(x)`))   # special care for data from Sweden around age 7-8
```



# PCA  and SVD over log-mortality tables


- [ ] facet screeplots for gender and countries
- [ ] comment the screeplots
- [ ] comment the correlation circles
- [ ] comment the individual plots

In the next chunks we compute the PCA (with standardization and centering) for all Countries and
Genders in the database. The `dplyr` pipeline prepares grouped tibble corresponding
to the different countries and genders. The collection is fed to  `group_map`
which attempts to compute pca for each group. This output is
a list of (key, value) pairs.

Henceforth,  we exclude data from Sweden since for several years after $2000$, 
mortality risks at some ages between $5$ and $10$ are zero. I do not know 
how reliable this information is. It might just be the outcome of random fluctuations: 
a modern Swedish generation typically comprises $110000$ children. Data 
obtained from larger (and possibly less healthy) European countries suggest 
that mortality risks between ages $5$ and $10$ are less than $5\times 10^{-5}$. 
This suggests that the yearly number of girls or boys dying ate age $x$ 
between $5$ and $10$ in Sweeden is dominated by a Poisson distribution with parameter $3$.
The probability of outcome $0$ under Poisson distribution with parameter $3$ is approximately `r round(exp(-3),2)`.



```{r}
#| label: nested-lt
#| appendix: true
#
#
# Nest the life tables data frame acording to Area, Country, Gender
lt_nested <- life_table  |>  
  filter(Age < 90)  |>       # no data above age 90 for some years in Spain
  nest(.by=c(Area, Country, Gender))  |>  
  mutate(data_tf=map(data, 
                  \(df)  pivot_wider(df, id_cols=Year, names_from=Age, values_from=`q(x)`)))
```


```{r}
#| label: taylored_prcomp
#| appendix: true
#
#
# Wrapper for prcomp for per Country/Gender PCA
prcomp_lt <- function(df, ...){
  df |> 
    tibble::column_to_rownames("Year") |>
    mutate(across(everything(), \(x) log(x, 10))) |> 
    scale() |> 
    prcomp(...)
}
```

```{r}
#| label: pca_s
#| appendix: true
#
#
# Add resultof  prcomp for per Country/Gender PCA as a list column

lt_nested <- lt_nested |> 
  mutate(pca = map(data_tf, \(x) try(prcomp_lt(x, rank=5)))) 
```





## Screeplots

The screeplots for standardized-centered PCAs for countries and genders shows

- the inertia projected on the first two principal components accounts for no less than $90\%$ of total inertia.
- for each country the inertia projected on the first two principal components is larger for Females than for Males
- The approximation by rank two matrices is worst for the Dutch matrices



```{r}
#| label: prepare_screeplots
#| appendix: true
#
#
# Add preparation for screeplots per Country/Gender PCA as a list column

lt_nested <- lt_nested |> 
  mutate(scrp = map(pca, \(x) drop_na(tidy(x, matrix="d"))))  
```

```{r}
#| label: df_screeplots
#| appendix: true
# 
#
# Add screeplots per Country/Gender PCA as a list column

lt_nested  |>  
  select(Country, Gender, scrp)  |>  
  unnest(cols=c(scrp)) |> 
  ggplot() +
    aes(x=PC, y=percent, fill=Country) +
    geom_col(position="dodge") +
    facet_wrap(~ Gender) +
    labs(
      title="Screeplots of log mortality quotients at Age < 90, Years between 1950-2020"
    )
```



But for all  countries and genders, most of the inertia is captured by the first two axes. 
USA and Canada exhibit a larger inertia on the second axis.


::: {.callout-note}

This suggests that for all countries,  genders, the scaled and centered log-quotient mortality matrices are well-approximated 
by rank-one matrices. However, recall that we measure approximation error using the Frobenius norm. We weight the different ages (columns) in a uniform way. This does not take into account the impact the different columns on the Life Expectancy.  

:::










Screeplots from PCA without standardization tell us in advance how well Lee-Carter 
approximation will perform for a given population.  The larger the share of 
inertia carried by the first component, the better the Lee-Carter approximation. 
From the figure above, we can infer that Lee-Carter approximation will be better 
for Female populations and than for Male populations. 

## Correlation circles 

For a given PCA, the correlation circle tells us  how close the different variables are from the plane spanned by the first two components  and where the projection of the variables lay on this plan. This can be the basis  of  some variables clustering procedures. 



The correlation circle delivered by `ade4` is not readable:
  - all variables are aligned with the first component as could have been predicted from the screeplot
  - labels overlapping hides a possible structure





```{r}
#| label: prep-co-circle
#| appendix: true
#
#
# Add preparation for correlation circles per Country/Gender PCA as a list column

prep_co_circle <- function(pco) {
  r <- with(pco,  t(t(rotation) * sdev[1:ncol(rotation)]))
  as_tibble(r) |> 
    rename_with(.fn = \(x) gsub('PC', '', x), .cols=everything()) |>
    mutate(row_id=rownames(r))
}
```




```{r}
#| appendix: true
#| label: plot_co_circle
#
#
#
plot_co_circle <- function(df, tit){
  df |> 
    ggplot() +
    aes(x=`1`, y=`2`, label=row_id) +
    geom_segment(
      aes(xend=0, yend=0, color=cut(
        as.integer(row_id),breaks=c(0, 6, 15, 65, 90), right=F))# ,
    ) +
  scale_color_viridis_d(name="Age") +
  ggrepel::geom_text_repel() +
  coord_fixed() +
  xlim(c(-1.1, 1.1)) + ylim(c(-1.1, 1.1)) +
  ggforce::geom_circle(aes(x0=0, y0=0, r=1), linetype="dashed") +
  theme_minimal() +
  labs(title=tit)
}

```


```{r}
#| label: build_plots_co_circle
#| appendix: true
#
#  
# Add correlation  circle plots per Country/Gender PCA as a list column

lt_nested <- lt_nested |>  
  mutate(cocirc= map(pca, prep_co_circle))  |> 
  mutate(tit=glue::glue("Cor circle {Country} {Gender}")) |> 
  mutate(p_co_circ=map2(cocirc, tit, plot_co_circle))
```

```{r}
#| appendix: true
#| label: plot-co-circle-canada
#
#
# Correlation circles for Canada
tmp <- lt_nested |>  
  filter(Country=='CAN') |> pull(p_co_circ)

tmp[[1]] + tmp[[2]]
```

For Canada


```{r}
#| appendix: true
#| label: plot-co-circle-italy
#
#
# Correlation circles for Italy

tmp <- lt_nested |>  
  filter(Country=='ITA') |> pull(p_co_circ)

tmp[[1]] + tmp[[2]]
```

For France and Italy,  as could have been guessed from screeplots, all centered and standardized
variables are highly correlated with the first component. This is even more striking for women than for men.


```{r}
#| appendix: true
#| label: plot-co-circle-usa
#
#
# Correlation circles for USA
tmp <- lt_nested |>  
  filter(Country=='USA') |> pull(p_co_circ)

tmp[[1]] + tmp[[2]]
```


For the Netherlands, Spain and the US, especially for men, variables  associated with old ages
are substantially correlated with the second component.

A common pattern emerges across countries and genders. Recall that PCA (and SVD) are uniquely defined up to signs of eigenvectors (singular vectors). The following description is taken from the correlation circle for women in France.

- Age groups `60-79` and `80-99` are packed in the same region.  The two age groups are almost aligned along parallel curves and  ranked according to age along each curve. The two curves look
oriented into opposite directions
- Age group `40-59` is also packed in a small region and almost ordered along the second component
- Age group `20-39` forms a SE-NW oriented cloud,  points are ordered according to ages
- Age group `0-19` spans the whole plot. Ages 0-7  are ordered around the second component. Ages `8-15` tend to cluster in the same region. Age `18` almost look like an outlier. Ages  `16-17`  are at an intermediate position between the outlier `18` and the `10-15` group.


Looking at the variables on the plane generated by the second and third principal components is also interesting.

- Ages below 20 lie around a SW-NE axis
- Ages between 20 and 40 lie on the south-west
- Ages between 40 and 60 are clustered and almost aligned on the sout-east
- Ages above 60 are almost aligned along a line
- Ages above 80 are almost aligned along another line on the north east quarter


## Individuals (Years)

If we plot the rows (years) on the plane spanned by the first two principal axes, patterns across
genders and countries again look very similar. Axis 1 roughly corresponds to a time axis. If we look
at points corresponding to the first 15 years, plots for European countries are very much alike.
For Female plots, the four points corresponding to years 1948-1951 follow the same paths in all
six European countries; for the US the path is different. For England and Wales, France and Italy,
the projections on the planed spanned by the first two principal axes look strikingly similar.



```{r}
#| appendix: true
#| label: plot-individuals
#
#
# Add row plots as a list column
lt_nested  <- lt_nested |>  
  mutate(rowplots=map2(pca, data_tf, \(pco, df2) autoplot(pco, df2))) |> 
  mutate(rowplots=map(rowplots, \(p) p + coord_fixed()  + ggrepel::geom_text_repel(aes(label=rownames)))) |> 
  mutate(rowplots=map2(rowplots, tit, \(x,y) x + labs(title=y)))
```



```{r}
#| appendix: true
#| label: row-plot-fr
#
#
# Row plot for USA
tmp <- lt_nested |>  
  filter(Country=='FRA') |> pull(rowplots)

tmp[[1]] + tmp[[2]]
```

# Canonical Correlation Analysis

Build a function that takes as input

+ a dataframe like `life_table_pivot`,
+ a couple  of countries, say `Spain` and `Sweeden`,
+ a vector of `Year`, say `1948:1998`
+ a `Gender` say `Female`

returns a matrix called $Z$ with rows corresponding to `Year` and columns corresponding to couples `(Country, Age)`.


# Lee-Carter model for US mortality

During the last century, in the USA and in western Europe,
central death rates at all ages have exhibited a general decreasing trend.
This decreasing trend has not always been homogeneous across ages.

The Lee-Carter model has been designed to model and forecast the
evolution of the log-central death rates for the United States during the XXth century.

Let $A_{x,t}$ denote the log central death rate at age $x$ during year $t\in T$
for a given population (defined by Gender and Country).

The Lee-Carter model assumes that observed loagrithmic central death rates
are sampled according to the following model

$$A_{x,t} \sim_{\text{independent}} a_x + b_x \kappa_t + \epsilon_{x,t}$$

where $(a_x)_x, (b_x)_x$ and $(\kappa_t)_t$ are unknown vectors that satisfy

$$a_x = \frac{1}{|T|}\sum_{t \in T} A_{x,t}\qquad \sum_{t\in T} \kappa_t = 0 \qquad \sum_{x} b_x^2 =1$$

and $\epsilon_{x,t}$ are i.i.d Gaussian random variables.

## US data

- [ ] Fit a Lee-Carter model on the American data (for Male and Female data) training on years `1933` up to `1995`.
- [ ] Compare the fit provided by the Lee-Carter model with the fit provided by a rank $2$
truncated SVD
- [ ] Compare vectors avec $(a_x)_x, (b_x)_x$ and $(\kappa_t)_t$ with appropriate singular vectors. 
- [ ] Use the Lee-Carter model to predict the central death rates for years $2000$ up to $2015$
- [ ] Plot predictions and observations for years $2000, 2005, 2010, 2015$


## Application of Lee-Carter model to a European Country


- [ ]  Fit a Lee-Carter model to a European country (Framce) using `demography::lca`
- [ ]  Comment
- [ ]  Compare with rank-2 truncated SVD
- [ ]  Use the Lee-Carter model to predict the central death rates for years $2000$ up to $2015$
- [ ]  Plot predictions and observations for years $2000, 2005, 2010, 2015$




```{r}
#| label: lca-fr
#| appendix: true
#
#
# Lee-Carter analysis
res <- demography::lca(
  demography::fr.mort, 
  ages=seq(20, 80), 
  years=seq(1948, 1998), 
  restype="logrates")
```


```{r}
#| label: lca-fr-plot
#| appendix: true
#
#
# Lee-Carter analysis plots
plot(res)
```




## Predictions of life expectancies at different ages

- [ ] Use Lee-Carter approximation to approximate residual life expectations
- [ ]  Compare with observed residual life expectations



::: {.callout-note title="References"}

- Lee, R. D., & Carter, L. R. (1992). *Modeling and Forecasting U. S. Mortality.* Journal of the American Statistical Association, 87(419), 659–671. https://doi.org/10.2307/2290201

- Preston, S. H., Heuveline, P., & Guillot, M. (2001). Measuring and Modeling Population Processes, Oxford: Blackwell Publishers, 2001, xv+ 291 pp.

:::




::::: {.content-visible unles-profile="solution"}

# Appendix 


```{r}
#| label: chunk-make-appendix
#| ref.label: !expr knitr::all_labels(appendix==TRUE)
#| echo: true
#| eval: false
#| code-fold: false
```


:::::