---
title: "Lab: Hierarchical Clustering"
date: "`r Sys.time()`"

execute:
  echo: true
  eval: true
  collapse: false
  message: false
  warning: false
  code-fold: false

format:
  html:
    output-file: lab-hclust.html
  pdf:
    output-file: lab-hclust.pdf

engine: knitr
---



{{< include _preamble.qmd >}}

```{r}
#| echo: true
#| message: false
#| code-fold: true
stopifnot(
  require(gt),
  require(skimr),
  require(GGally),
  require(patchwork),
  require(ggforce),
  require(glue),
  require(ggfortify),
  require(ggvoronoi),
  require(magrittr),
  require(broom),
  require(ggdendro),
  require(dendextend),
  require(plotly),
  require(tidyverse)
)

tidymodels::tidymodels_prefer(quiet = TRUE)

old_theme <-theme_set(
  theme_minimal(base_size=9, 
                base_family = "Helvetica")
)
```

```{r}
#| echo: true
#| code-fold: true
#| 
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  comment=NA,
  prompt=FALSE,
  cache=FALSE,
  echo=TRUE,
  results='asis'
)
```


```{r}
#| echo: true
#| code-fold: true
#| 
gc <- options(ggplot2.discrete.colour="viridis")
gc <- options(ggplot2.discrete.fill="viridis")
gc <- options(ggplot2.continuous.fill="viridis")
gc <- options(ggplot2.continuous.colour="viridis")
```

Preamble
---------

Hierarchical clustering builds *dendrograms*

Explore the data structure: dendrograms (objects of class `dendrogram`) are represented by *lists of lists  with attributes* (not by `tibbles`). 

The dendrograms created from objects of class `hclust` represent *planar binary trees*. 

::: {.callout-note}

### Question

- How do you define abstractly planar binary trees?
- In dendrograms created from objects of class `hclust`, what do the leaf nodes represent? 
- In dendrograms created from objects of class `hclust`, what do the internal nodes represent ?

:::

::: {.callout-tip}

Keep an eye on  [Introduction to dendextend](https://cran.r-project.org/web/packages/dendextend/vignettes/dendextend.html#a-dendrogram-is-a-nested-list-of-lists-with-attributes) by the package author Tal Galili. 


:::


Playing with a toy dendrogram
-----------------------------


```{r}
dend <- 1:5 %>% 
  dist %>% 
  hclust(method="ward.D2") %>% 
  as.dendrogram
```

```{r}
dend |>  plot()
```

```{r}
dend  |>  head(10)
```

::: {.callout-note title="Question"}

- List the generic functions with registered methods for class `dendrogram`

:::

::: {.content-visible when-format="solution"}

::: {.callout-tip title="Solution"}

```{r}
#| label: methods-dendrogram

sloop::s3_methods_class("dendrogram") |> 
  select(generic) |> 
  gt::gt() |> 
  gt::tab_caption("Fonctions génériques pour `dendrogram`")
```
:::

:::

::: {.callout-note title="Question"}

Explain attribute `members`  in the next chunk.

Hint: Nodes are identified by their *prefix order index*. 

:::

```{r}
dend |>  
  get_nodes_attr("members", 
                 id = c(1, 2, 5, 7)) 
```

::: {.content-visible when-profile="solution"}

::: {.callout-tip title="Solution"}

The dendrogram is traversed using depth-first-search. For each node, attribute `members` is the number of leaves 
in the subtree rooted at that node. 

We start form the root with 5 leaves/members, then proceed to left child with 2 leaves/members .... 

:::

:::





::: {.callout-note title="Question"}

- What does `labels()` return?
- What does function `rotate` do ?

:::

```{r}
dend |>  labels()

dend |> 
  rotate(c(1,2,4,5,3)) |> 
  plot()
```

::: {.content-visible when-profile="solution"}

::: {.callout-tip title="Solution"}

`labels()` returns leaves labels ordered by depth-first-search tarversal.

:::

:::

::: {.callout-note type="Question"}

What is the *height* of a branch?

Just look at what happens when `hclust` uses method `ward.D2`

:::  


::: {.content-visible when-profile="solution"}

::: {.callout-tip title="Solution"}

```{r}
hc9 <- 1:9  |>  dist()  |>  hclust(method="ward.D2")

hc9$height

dend9 <- as.dendrogram(hc9)

dend9 |>  plot()
#get_branches_heights(dend9)

hc9$height^2
```

This is consistent with the `ward.D2` clost  for merging clusters with $n_A, n_B$ elements and centroids $\mu_A, \mu_B$

$$\left(2 \frac{n_An_B}{n_A+n_B} \|\mu_A - \mu_B\|^2\right)^{1/2}$$


:::

:::

::: {.callout-note type="Question"}

What is a cophenetic distance ?

:::

```{r}
cophenetic(rotate(dend, c(1,2,4,5,3)))
cophenetic(dend)
```

::: {.content-visible when-profile="solution"}

::: {.callout-tip title="Solution"}

Just have a look at `?cophenetic`

:::

:::









::: {.callout-note title="Question"}

How is attributed `height` computed? What is its purpose? 

What kind of tree traversal is used by `get_nodes_...` helpers?

:::


```{r}
#| label: Get-various-attributes
dend |> 
  rotate(c(1,2,4,5,3)) |> 
  get_nodes_attr("height") 
```
  

```{r}
dend %>% 
  get_nodes_attr("members")

```

::: {.content-hidden unless-profile="TOUT"}

Tweaking a dendrogram
-----------------------

Why should we do that? 


How should we do that? 

```{r}

```

:::

USArrests
---------

We work on `USArrests` dataset. We want to classify the 50 (united) states on the basis of the arrests profile and the urbanization rate. We rely on hierarchical, bottom-up classification.

```{r}
data("USArrests")

USArrests <- USArrests %>% 
    tibble::rownames_to_column(var="region")

USArrests <- USArrests %>%
    mutate(region = tolower(region))

rownames(USArrests) <- USArrests$region

glimpse(USArrests)
```

```{r}
source("./UTILS/make_biotiful.R")
```

```{r}
make_biotifoul(USArrests, .f=is.numeric)
```

The function `dist` is used to calculate pairwise distances between individuals. 

::: {.callout-note}

### Question 

Compute pairwise distances between rows of `USArrests` (with and without scaling)

:::

::::: {.content-visible when-profile="solution"}  


```{r}
dist.1 <-  USArrests %>% 
    select(where(is.numeric)) %>% 
    dist()
```

```{r}
dist.2 <- USArrests %>% 
    select(where(is.numeric)) %>% 
    scale %>% 
    dist()
```

:::::

::: {.callout-note}

### Question

Perform hierarchical clustering on *unscaled* and *scaled* dataset.

:::

::::: {.content-visible when-profile="solution"}  
::: {.callout-tip title="Solution"}

```{r}
hcl.1 <- hclust(dist.1, method = "ward.D2")
hcl.2 <- hclust(dist.2, method = "ward.D2") # scaled
```

:::
:::::

::: {.callout-note}

### Question

Draw a choropleth map of the US. Color states according to cluster assignment.   

Hint: use `ggplot2::map_data`

:::



::::: {.content-visible when-profile="solution"}  
::: {.callout-tip title="Solution"}


```{r}
mutate(USArrests, 
       .cluster = factor(cutree(hcl.1, 5))) %>% 
  inner_join(map_data("state"), by = "region") %>%
  ggplot() +
  aes(x=long, y=lat, group=region, fill=.cluster) +
  geom_polygon() +
  scale_fill_viridis_d() +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  ggtitle("Components of arrest data") +
  theme(legend.position = "none")
```

:::
:::::


The `dendrogram` class
----------------------

::: {.callout-note}

### Question

Exploration of results of hierarchical clustering (objects of class `hclust`) is facilitated by converting to class `dendrogram`.

:::

::::: {.content-visible when-profile="solution"}  

::: {.callout-tip title="Solution"}

```{r}
dend.1 <- hcl.1 %>% 
  as.dendrogram()

length(dend.1[[1]][[2]])

```


```{r}
class(dend.1)
class(unclass(dend.1))

# methods(class=class(dend.1)) |> head()
```

```{r}
dend.1 |> head()
```

:::

:::::

::: {.callout-note title="Question"}

:::


::::: {.content-visible when-profile="solution"}  


::: {.callout-tip title="Solution"}

```{r}
dend.1 %>% 
  ggdendrogram(rotate = TRUE,labels = T) +
  ggtitle("Dendrogram for USArrests") +
  ggdendro::theme_dendro() +
  scale_y_reverse(expand = c(0.2, 0))
```  

:::

:::::

::: {.callout-note}

### Question



:::
  
::::: {.content-visible when-profile="solution"}  


::: {.callout-tip title="Solution"}

```{r}
# label(dend.1)

dend.2 <-  as.dendrogram(hcl.1)
# order it the closest we can to the order of the observations:
dend.2 <- dendextend::rotate(dend.2, 1:50)
# Color the branches based on the clusters:
dend.2 <- color_branches(dend.2, k=3) #, groupLabels=iris_species)
# Manually match the labels, as much as possible, to the real classification of the flowers:
# labels_colors(dend.2) <-
#    rainbow_hcl(3)[sort_levels_values(
#       as.numeric(iris[,5])[order.dendrogram(dend.2)]
#    )]
```

:::

:::::  


Ward method 
-----------


The `meth=ward.D2` option allows you to aggregate individuals according to the method of Ward, that is, according to the variance.

::: {.callout-note}

### Question

What is the distance used? Describe the method of *classification by variance*?

:::

::::: {.content-visible when-profile="solution"}  

The output `clas$height` gives the jump height of the dendrogram to
each new iteration. In the case of Ward's method, she
is proportional to the loss of inter-class variance.


:::::

::: {.callout-note}

### Question

1. How many groups are there at step 0? at the last step?

2.  How many iterations are there?

3.  Recall the definition of inter-class variance.

4.  What is the inter-class variance at step 0? at the last
step? How is it going according to the number of groups
(or according to the number of iterations)?

5.  By comparing the total inertia and the `clas$height' output,
find the coefficient of proportionality between the loss of
inter-class variance and height of jumps.


:::


Choice of the number of classes
-------------------------------

::: {.callout-note}

### Question

1.  Plot the curve corresponding to the loss of variance inter in
as a function of the number of iterations :

2.  Select the "optimal" number of classes.

3.  Verify that, for the number of classes chosen, the number by
class is sufficient (we can use the `cutree` function).

4.  These classes can be represented using a dendrogram

5.  You can also colour the leaves of the tree corresponding to
 a class. To do this, install and load the package `dendextend'.


:::



Link with PCA 
-------------


We will represent the classes obtained in the
factorial design(s) obtained by the PCA. This will make it possible to represent the classes and describe them according to the variables initials.


::: {.callout-note}

### Question

Represent the coordinates of the individuals in each group in
the first factorial plane (with one color for each class).
The vector generated by `cutree' can be used to form a
color vector. Interpretation.


:::



::::: {.content-visible when-profile="solution"}  


```{r}
pca_arrests <- USArrests |> 
  select(where(is.numeric), -UrbanPop) |> 
  scale() |> 
  prcomp()

hcl_pca <- pca_arrests |>  
  augment() |> 
  select(.fittedPC1, .fittedPC2) |> 
  dist() |> 
  hclust(method="ward.D2")


```



```{r}
mutate(USArrests, 
       .cluster = factor(cutree(hcl_pca, 3))) %>% 
  inner_join(map_data("state"), by = "region") %>%
  ggplot() +
  aes(x=long, y=lat, group=region, fill=.cluster) +
  geom_polygon() +
  scale_fill_viridis_d() +
  ggtitle("Components of USArrest data", "Clustering after PCA") +
  theme(legend.position = "none")
```



```{r}
p1 <- autoplot(pca_arrests, data=USArrests, loadings=T, loadings.label=T) +
  aes(x=PC1, y=PC2, color=Murder, label="region")+
  geom_point() +
#  ggrepel::geom_text_repel()  +
  coord_fixed() +
  theme(legend.position="none")

```

```{r}
p2 <- autoplot(pca_arrests, data=USArrests) +
  aes(x=PC1, y=PC2, color=Murder, label=region)+
  geom_point() +
  ggrepel::geom_text_repel()  +
  coord_fixed()
```

```{r}
p1 + p2
```
```{r}
(
mutate(USArrests, 
       .cluster = factor(cutree(hcl_pca, 3))) %>% 
  inner_join(map_data("state"), by = "region")  %>%
  ggplot() +
  aes(x=long, 
      y=lat, 
      label=factor(region), 
      fill = factor(.cluster)) +
  geom_polygon() +
  scale_fill_viridis_d() +
  ggtitle("USA") +
  theme(legend.position = "none")
  ) |> 
    ggplotly()
```

:::::


Cophenetic distance
-------------------

::: {.callout-note}

### Question

:::



::::: {.content-visible when-profile="solution"}  

```{r}
dist.coph.1 <- cophenetic(dend.1)
```


:::::


Cophenetic distance between dendrograms 
----------------------------------------


::::: {.content-visible when-profile="solution"}  



```{r}
data("iris")

hcl.iris <- iris %>% 
  select(where(is.numeric)) %>% 
  scale() %>% 
  dist() %>% 
  hclust(meth="ward.D2")

dend.iris <-  dendro_data(hcl.iris)

dend.iris %$% (
  ggplot() +
  geom_segment(data = segments,
               aes(x = x, y = y,
                   xend = xend, yend = yend)
  ) +
  geom_text(data = labels,
            aes(x = x, y = y,
                label = label, hjust = 0),
            size = 3
  ) +
  coord_flip() +
  scale_y_reverse(expand = c(0.2, 0))
)
  
```
:::::




References
--------------

[`ggdendro`](https://andrie.github.io/ggdendro/)

[`dendroextra`](https://github.com/jefferis/dendroextras)

[`hier_clust`, `tidyclust`](https://tidyclust.tidymodels.org/articles/hier_clust.html)
