---
date: "`r Sys.time()`"
title: "Lab: Life Tables (1948-2016)"


execute:
  echo: false
  eval: true
  collapse: true
  warning: false
  message: false

format:
  html:
    output-file: lab-lt-1948-2016.html
  pdf:
    output-file: lab-lt-1948-2016.pdf


engine: knitr

---


{{< include _preamble.qmd >}}


```{r}
#| echo: false
#| include: false
#| message: false
#| warning: false

stopifnot(
  require(patchwork),
#  require(httr),
  require(arrow),
  require(glue),
  require(ineq),
  require(here),
  require(skimr),
  require(stringr),
  require(magrittr),
  require(plotly),
#  require(slider),
  require(gt),
  require(rlang),
  require(tidyverse)
)

old_theme <- theme_set(theme_minimal())
```

# Introduction

Life tables 

- [John Gaunt. The first life table. 1663](https://doi.org/10.1098/rsnr.1938.0017) 
- [De Witt. Valuation of life annuities. 1671](https://mathshistory.st-andrews.ac.uk/Projects/Daxenberger/chapter-7/)
- [Actuarial science](https://en.wikipedia.org/wiki/Actuarial_science)


> A  period life table is supposed to represent the mortality conditions at a specific moment in time. Period life tables are said to be synthetic in that each age group of data comes from a different birth cohort.



Demography 

- History
- Perspective


# Getting the data 

::: {.callout-note}

[http://www.mortality.org](http://www.mortality.org) 

[http://www.lifetable.de](http://www.lifetable.de)

:::

# First attempt at understanding


[Methodology](https://www.lifetable.de/File/GetDocument/methodology.pdf)


## Glossary 

## Connections between terms


# Data wrangling 

```{r}
#| label: data-path
#| appendix: true
res_path <- fs::path('~/Dropbox/HMD/res') |> fs::path_expand()
res_url <- 'https://www.lifetable.de/File/GetDocument/data/hld.zip'
dest_path <- fs::path_join(c(fs::path_dir(res_path), fs::path_file(res_url)))
dpath <- fs::path('~/Dropbox/HMD/res') |>  fs::path_expand()

if (!fs::file_exists(res_path)) {
  download.file(res_url, destfile = dest_path)
  zip::unzip(dest_path)
}
```

```{r}
#| label: res-col-spec
#| appendix: true 
res_col_spec <- cols(
  Country = col_character(),
  Region = col_character(),
  Residence = col_character(),
  Ethnicity = col_character(),
  SocDem = col_character(),
  Version = col_character(),
  `Ref-ID` = col_character(),
  Year1 = col_integer(),
  Year2 = col_integer(),
  TypeLT = col_integer(),
  Sex = col_integer(),
  Age = col_integer(),
  AgeInt = col_double(),
  `m(x)` = col_double(),
  `q(x)` = col_double(),
  `l(x)` = col_double(),
  `d(x)` = col_double(),
  `L(x)` = col_double(),
  `T(x)` = col_double(),
  `e(x)` = col_double(),
  `e(x)Orig` = col_character()
)
```


```{r}
#| label: load-csv-file
tb <- readr::read_csv(dpath, col_types = res_col_spec)
```


```{r}
# problems(tb)
```

```{r}
tb |> glimpse()
```

```{r}
#| label: save-partitioned-parquet
#| appendix: true

ds_path <- '~/Dropbox/HMD/hld-part.parquet'

if (!fs::file_exists(ds_path)){
  tb |> 
    group_by(Country, Sex) |>
    arrow::write_dataset(path=ds_path)
}
```

```{r}
#| label: country-codes
#| appendix: true

country_iso3 <- c(
  "France"="FRA",
  "Canada"="CAN",
  "Netherlands"="NLD",
  "Sweden"="SWE",
  "Italy"="ITA",
  "Spain"="ESP",
  "Portugal"="PRT",
  "USA"="USA"
)

tb_iso3 <- tibble(
  name=names(country_iso3), iso3=country_iso3
)
```


```{r}
#| label: col-selection
#| appendix: true

ds <- open_dataset(ds_path)

ds$schema
```

```{r}
#| label: load-lt-countries
tb <- ds |>
  filter(
    Country %in% country_iso3, 
    Sex %in% c(1L, 2L), 
    Region=='0', 
    Residence=='0',
    SocDem=='0',
    Version=='1',
    Year1 >= 1948, 
    Ethnicity=='0', 
    AgeInt==1,
    TypeLT==1) |> 
    select(-c(Region, Ethnicity,Year2,AgeInt, TypeLT, Residence, SocDem, Version)) |> 
  collect()
```

```{r}
#| label: crude-averaging
#| appendix: true
tb <- tb  |>  
  group_by(Country, Sex, Year1, Age)  |>  
  summarise(across(ends_with('(x)'), mean)) |> 
  ungroup()
```


# Saving and reloading

# The evolution of mortality quotients in Western Europe and the USA since WWII

## Facets

## Animations

## Focus on infant mortality

## Focus on young adults

## Older people 


========================================================================================

```{r coding_tables}
#| echo: false
#| eval: false
country_code <- list(fr_t='FRATNP',
                     fr_c='FRACNP',
                     be='BEL',
                     gb_t='GBRTENW',
                     gb_c='GBRCENW',
                     nl='NLD',
                     it='ITA',
                     swe='SWE',
                     sp='ESP',
                     us='USA')

countries <- c('fr_t', 'gb_t', 'nl', 'it', 'sp', 'swe', 'us')

country_names <- list(fr_t='France',     # total population
                     fr_c='France',      # civilian population
                     be='Belgium',
                     gb_t='England & Wales',    # total population
                     gb_c='England & Wales',    # civilian population
                     nl='Netherlands',
                     it='Italy',
                     swe='Sweden',
                     sp='Spain',
                     us='USA')

gender_names <- list('b'='Both',
                     'f'='Female',
                     'm'='Male')
```


```{r}
#| echo: false
params <- list(
  year_e=2020, 
  year_p=1950
)
```



```{r}
#| eval: true
#| echo: true
life_tables <- tb |>
  mutate(Iso3 = factor(Country)) |>
  mutate(Gender = factor(Sex, labels=c("Male", "Female"))) |>
  mutate(Area = fct_collapse(Iso3, 
    SE = c("PRT", "ESP", "ITA", "FRA"), 
    NE = c("NLD", "SWE", "CAN"), 
    USA = "USA")) |> 
  rename(Year=Year1) |> 
  select(Country, Gender, Year, Age, `q(x)`) 
```



# Definitions

Check on [http://www.mortality.org](http://www.mortality.org) the meaning of the different columns. 

See: *Demography: Measuring and Modeling Population Processes* by 
SH Preston, P Heuveline, and  M Guillot. Blackwell. Oxford. 2001.

Document [Tables de mortalité françaises pour les XIXe et XXe siècles et projections pour le XXIe siècle](https://www.lifetable.de/data/FRA/FRA000018061997CY1.pdf) contains
detailed information on the construction of Life Tables for France.

## Period tables versus cohort tables 

Two kinds of Life Tables can be distinguished: *Period tables* (*Table du moment*) which
contain for each period (here a period is a calendar year), the mortality risks at different age ranges (here, we have one year ranges) for that very period; and *Cohort tables* (*Tables de génération*) which contain for a given birthyear, the   mortality risks at which an individual born during that year has been exposed. 

The life tables investigated in this lab are *Period table* (*Table du moment*). According to the cited documents  ([Methodology](https://www.lifetable.de/File/GetDocument/methodology.pdf), Vallin and Meslé), building life tables  requires decisions and doctoring (imputations, smoothing, ...).

## Lexis diagrams 

Lexis diagrams provide a graphical device that summarizes the construction of mortality  quotients (and other rates in demography). 


```{r}
#| label: lexis-diagram
#| code-fold: true
#| appendix: true
birth_dates <- lubridate::as_date("1999-01-01") + lubridate::duration(sample(2*365, size=20, replace=T),units="day")
death_dates <- lubridate::as_date("2009-07-01") + lubridate::duration(sample(3*365, size=20, replace=T),units="day")

b_period <- lubridate::as_date("2010-01-01")
b_frame <-  lubridate::as_date(b_period - lubridate::duration(1, units = "year"))
b_age <- 10L

tb_ld <- tibble(birth=birth_dates, death=death_dates)
```

```{r}
#| label: lexis-diagram-2
#| fig_title: "Lexis diagram"
#| appendix: true
tb_ld |> 
  ggplot() +
  geom_segment(aes(x=b_frame,
                   xend=death,
                   y=lubridate::interval(birth, b_frame)/lubridate::years(1),
                   yend=lubridate::interval(birth, death)/lubridate::years(1))
               ) +
  annotate(geom="rect",
           xmin=b_period,
           xmax=b_period + lubridate::duration(1, units = "year"),
           ymin=b_age,
           ymax=b_age + 1L,
           fill="grey",
          alpha=.5) +
  ylab("Age") +
  xlab("Time") +
  coord_cartesian(xlim=c(lubridate::as_date(b_period - lubridate::duration(6, units = "months")), 
                         lubridate::as_date(b_period + lubridate::duration(18, units = "months"))),
                  ylim=c(b_age - .5, b_age+1.5)) +
  labs(
    title="A Lexis diagram",
    subtitle = "for mortality quotient at Age 10 during Year 2010-11"
  ) +
  theme_minimal()
```

Each line represents the *life line* of an individual born during years 1999 and 2000 and deceased beetween mid 2009 and mid 2012. In order to compute the mortality quotient at age 10 for year 2010, we have to compute the relevant *number of occurrences*, that is the number of segments ending in the grey rectangle, and the *sum of exposure times*, which is proportional to the sum of the lengths of the segments crossing the grey rectangle. 

Have a look at [Lexis diagram](https://en.wikipedia.org/wiki/Lexis_diagram), at Preston *et al*, at []().

Definitions can be obtained from [www.lifeexpectancy.org](http://www.lifeexpectancy.org/lifetable.shtml). We translate it into mathematical (rather than demographic) language. 

The mortality quotients define a probability distribution over $\mathbb{N}$. This probability distribution is a _construction_ that reflects the *health situation* in a population at a given time. This probability distribution does not describe the sequence of health situations experienced by a _cohort_ (people born during a specific year).

> One  works with a period, or current, life table (_table du moment_). This summarizes the mortality experience of persons across all ages in a short period, typically one year or three years. More precisely, the death probabilities $q_x$ for every age $x$ are computed for that short period, often using *census information* gathered at regular intervals. These $q_x$'s are then applied to a hypothetical cohort of $100 000$ people over their life span to produce a life table. 


::: {.callout-note}

Columns `q(x)` and `m(x)` (for mortality quotient  and central death rate) are displayed in scientific notation to stress the fact that their range extends over several orders of magnitude. 

:::

::::: {.content-visible when-profile="solution"}  

```{r}
#| label: french-lt-excerpt
#| appendix: true
#| 
tb |> 
    filter(Country=='FRA', Year1== 2010, Sex==2, Age<10|Age==64|between(Age, 80, 88)) |> 
    select(Age, `q(x)`, `m(x)`, `l(x)`, `d(x)`, `L(x)`, `T(x)`, `e(x)`) |> 
    gt::gt() |> 
    gt::tab_caption("Excerpt of French Period Life Table for Year 2010, Females") |> 
    gt::fmt_scientific(columns=c(`q(x)`, `m(x)`))
```
:::::


## Understanding the columns of the life table

In the sequel, we denote by $F_{t}$ the _cumulative distribution function_ for year $t$. We agree on $\overline{F}_t = 1 - F_t$ and $F_t(-1)=0$.  Henceforth, $\overline{F}$ is called the *survival* function.   





`q(x)`
: (age-specific) `risk of death` at age $x$, or `mortality quotient` at given age $x$ for given year $t$.

::: {.callout-note}

### About the definition of `q_{t,x}`

Defining and computing `q_{t,x}`  does not boil down to knowing the number of people at age $x$ at the beginning of ear $t$ and knowing how many of them died during year $t$. If we want to be rigorous, we need to know all life lines in the Lexis diagram, or equivalently, how many people at Age $x$ were alive on each  day of Year $t$.  That is, we need accurate *death counts* and *exposures to risk*.      

:::


::: {.callout-caution}

### Mortality quotients define a probability distribution

For a given year $t$, the sequence of mortality quotients define a  survival  function $\overline{F}_t$ using the following recursion: 

$$q_{t,x} = \frac{\overline{F}_t(x) - \overline{F}_t(x+1)}{\overline{F}_t(x)}$$
with boundary condition $\overline{F}_t(-1) =1$.

This recursion can also be read as: 

$$\overline{F}_{t}(x+1) = \overline{F}_{t}(x) \times (1-q_{t,x+1})\, .$$

Up to (non-trivial) corrections, this artificial probability distribution is used to define and compute life expectancies. 

:::

::: {.callout-note}

### {{< fa hand-point-right >}}

$q_{t,x}$ is the *hazard rate* of $\overline{F}_t$ at age $x$.

:::


`m(x)`
: *central death rate* at age $x$  during year $t$. This is connected with $q_{t,x}$ by
$$m_{t,x} = -\log(1- q_{t,x}) \,,$$ 

or equivalently 
$$q_{t,x} = 1 - \exp(-m_{t,x})$$

::: {.callout-note}

### About central death rate

If we want to define an absolutely continuous probability distribution $G$ over $[0,\infty)$ so that 
$G$ and $F$ coincide over integers and $G$ has piecewise constant hazard rate, we can pick
$m_{t,x}$ as the piecewise constant hazard rate. 

:::



`l(x)`
: the so-called _survival function_: the scaled proportion of persons alive at age $x$. These values are computed recursively from the $q_{t,x}$ values using the formula 

$$l_t(x+1) = l_t(x) \times (1-q_{t,x}) \, ,$$
with $l_{t,0}$, the *radix* of the table, (arbitrarily) set to $100000$. In the table `lx` is rounded to the next integer


Function $l_{t,\cdot}$ and $\overline{F}_t$ are connected by 

$$l_{t,x + 1} = l_{t,0} \times \overline{F}_t(x)\,.$$

`dx`
: $d_{t,x} = q_{t,x} \times l_{t,x}$. 
The fictitious number of deaths occurring at age $x$ during year $t$. Again this is a rounded quantity. 

`T(x)`
: Total number of person-years lived by the cohort from age $x$ to $x+1$. This is the sum of the years lived by the $l_{t, x+1}$ persons who survive the interval, and the $d_{t,x}$ persons who die during the interval. The former contribute exactly $1$ year each, while the latter contribute, on average, approximately half a year, so that $L_{t,x} = l_{t,x+1} + 0.5 \times d_{t,x}$. This approximation assumes that deaths occur, on average, half way in the age interval x to x+1. Such is satisfactory except at age 0 and the oldest age, where other approximations are often used. 

Compare with the denominator in the definition of `q(x)` and its description using the Lexis diagram. 

*We will stick to a simplified vision  $L_{t,x}= l_{t,x+1}$*

`e(x)`:
: Residual Life Expectancy at age $x$ and year $t$

This is the  expectation of $X -x$ for a random variable $X$ distributed according to $\overline{F}_t$ conditionnally on the event  $\{ X \geq x \}$. That is $e_{t,x}$ is the expectation of the probability distribution defined by $\overline{F}_t(\cdot + x-1)/\overline{F}_t(x-1)$. 

::: {.callout-note}

### Question 

Check dependencies between columns

:::

::::: {.content-visible when-profile='solution'} 
 
 
```{r}
#| label: sanity-checks
#| eval: false
#| appendix: true
life_tables |> 
  filter( Year>=1948, Age < 90, Gender != "Both") |> 
  group_by(Country, Year, Gender) |> 
  summarise(m1 =max(abs(lx -dx -lead(lx))/lx, na.rm = T), 
            m2 =max(abs(lx * qx -dx)/dx, na.rm=T),
            m3 =max(abs(Lx -lx * (1 + qx * (ax-1)))/Lx, na.rm=T),
            m4 =max(abs(1-exp(-mx)-qx)/qx, na.rm=T),
            .groups = "drop") |> 
  select(Year, Country, Gender, m1, m2, m3, m4) |>  
  rename(lx=m1, dx=m2, Lx=m3, qx=m4) |> 
  group_by(Country, Gender) |> 
  slice_max(order_by = desc(qx), n = 1)  |> 
  ungroup() |> 
  gt() |> 
  tab_header(
    title = "Life table (relative discrepancies)",
    subtitle = ""
  ) |> 
  fmt_engineering(columns = ends_with("x"), 
                  decimals=2,
                  drop_trailing_zeros = T ) |> 
  tab_source_note(source_note = "From https://mortality.org")
  
```


:::::



# Western countries aroun 1950

Several pictures share a common canvas: 

::: {.callout-note}

### Question 

Plot mortality quotients (`q(x)`) against age using a logarithmic scale on the
$y$ axis. Countries are identified by aesthetics (`shape`, `color`, `linetype`). 

1. Use facetting to plot `q(x)` of all countries at all ages for years 1950, 1960, ...,  2010.
2. Use `plotly` to build an animated plot using `Year` for the `frame` aesthetics. 

:::


::: {.callout-tip}

Abiding to the DRY principle, define a prototype `ggplot`  (alternatively `plotly`) object. 

The prototype will then be fed with different datasets and decorated
and arranged for the different figures.

:::

::::: {.content-visible when-profile='solution'} 

```{r}
#| label: protogg
#| 
dummy_data <- filter(life_tables, FALSE)

proto_plot <- dummy_data |>
  ggplot() +
    aes(x=Age,
        y=`q(x)`,
        col=Country,
        linetype=Country) +
    scale_y_log10() +
    scale_x_continuous(breaks = c(seq(0, 100, 10), 109)) +
    ylab("Mortality quotients") +
    labs(linetype="Country") +
    theme_minimal()
```

:::::


::::: {.content-visible when-profile="solution"}  

```{r}
#| label: 65_years
#| message: false
#| warning: false
#| echo: false

p_1 <-  with(params,
  (proto_plot + (
    life_tables |>
    filter(Year %in% c(year_p, year_e), 
      Age<90,
      Gender != 'Both')) +
    geom_point(size=.1) +
    geom_line(linewidth=.2) +
    labs(title = glue("European countries and US, {year_p}-{year_e}"),
    subtitle = "Sweden and the Netherlands consistently ahead") +
    facet_grid(rows=vars(Year), cols=vars(Gender))
  )
) 
```  

::: {.content-hidden unless-format="html"}  

```{r}
p_1 |> 
    plotly::ggplotly()
```

:::

::: {.content-hidden when-format="html"}  

```{r}
p_1 
```

:::

:::::

::::: {.content-visible when-profile="solution"}  


```{r}
#| label: film_1948_2013
proto_plt2 <-
  ggplot() +
  aes(x=Age, y=`q(x)`, colour=Country, frame=Year, linetype=Country) +
  geom_point(size=.1) +
  scale_y_log10() +
  scale_x_continuous(breaks = c(seq(0, 100, 20), 109)) +
  xlab("Age") +
  ylab("Mortality quotients") +
  facet_grid(cols=vars(Gender)) +
  theme_minimal()
```

```{r}
#| label: film_1948_2013_plotly
p_2 <- with(params,
  (proto_plt2 +
    (life_tables |>
      filter(between(Year, year_p, year_e), 
             Year %% 10 == 0, 
             Gender != 'Both', 
             Age < 90))  +
    ggtitle(glue::glue("Mortality quotient {params$year_p}-{params$year_e}: Europe catches up")))
)
```

::: {.content-hidden unless-format="html"}  

```{r}
p_2 |>
  plotly::ggplotly()

```

:::

In 1948, NE (Northern Europe) and the USA exhibit comparable mortality quotients at all ages for the two genders, the USA looking like a more dangerous place for young adults. Spain lags behind, Italy and France showing up at intermediate positions.

By year 1962, SE (Northern Europe)has almost caught up the USA. Italy and Spain still have higher infant mortality while mortality quotients in the USA and France are almost identical at all ages for both genders.
Mortality quotients attain a minimum around 10-12 for both genders. In Spain the minium
central death rate has been divided by almost ten between 1948 and 1962.

If we dig further we observe that the shape of the male mortality quotients curve changes over time.
In 1962, in the USA and France, mortality quotients exhibit a sharp increase between years 12 and 18,
then remain almost constant between 20 and 30 and afterwards increase again. This pattern shows up in other countries but in a less spectacular way.

Twenty years afterwards, during years 1980-1985, death rates at age 0 have decreased at around $1\%$
in all countries while it was $7\%$ in Spain in 1948. The male curve exhibits a plateau between ages 20 and 30. Mortality quotients at this age look higher in France and the USA.

By year 2000, France is back amongst European countries (at least with respect to mortality quotients). Young adult mortality rates are higher in the USA than in Europe. This phenomenon  became more pregnant during decade 2010-2020.

:::::


::: {.callout-note}

### Question

Plot ratios between mortality quotients (`q(x)`) in European countries and
mortality quotients in the USA in 1948.

:::

::::: {.content-visible when-profile='solution'} 
 
 
```{r}
#| label: build-eur-us-table
#| message: false
#| warning: false
#| echo: false

# simplified_life_table <- with(params, 
#   life_tables |>
#   filter(between(Year, year_p, year_e), Age<90, Gender!="Both") |>
#   select(Country, Year, Gender, Age, qx)
# )

eur_table <- life_tables |>
  filter(Country!='USA')

us_table <- life_tables |>
  filter(Country=='USA') |>
  select(-Country)

eur_us_table <-  eur_table |>
  inner_join(us_table, by=c('Age', 'Year', 'Gender')) |>
  mutate(Ratio=`q(x).x`/`q(x).y`)
```

```{r}
p <- with(params,
  (filter(eur_us_table, Year %% 10==0)  |>
    ggplot() +
    aes(x=Age,
        y=Ratio,
        frame=Year) +
    aes(linetype=Country, show.legend = FALSE) +
#    aes(color=Area, show.legend = FALSE) +
    scale_y_log10() +
    scale_x_continuous(breaks = c(seq(0, 100, 10), 109)) +
    geom_point(size=.1) +
    geom_smooth(method="loess",
                formula= 'y~ x',
                se=FALSE,
                span=.1,
                size=.1) +
    ylab("Ratio of mortality quotients with respect to US") +
    ggtitle(label = glue("European countries with respect to US,{year_p}-{year_e}"), 
            subtitle = "Sweden consistently ahead") +
   facet_grid(rows = vars(Gender))
  )) 

gp <- p |>
  ggplotly()

gp 
```


```{r}
for (i in seq_along(gp$x$data)) {
  gp$x$data[[i]]$showlegend <- FALSE
}

gp
```


```{r}
(p + with(params,
  filter(eur_us_table, Year %% 10==0, Age <= 10)) +
  geom_point(size=.5) ) |>
  ggplotly()
```

```{r}
(p + with(params,
  filter(eur_us_table, Year %% 10==0, between(Age, 15, 25))) ) |>
  ggplotly()
```



:::::


::: {.callout-note}

### Comment

This animation reveals less than the preceding one since we just have ratios
with respect to the USA. But the patterns followed by European societies emerge in a more transparent way.
The divide between northern and southern Europe at the onset of the period is even more visible.
The ratios are important across the continent: there is a factor of 10 between spanish and swedish
infant mortality rates. But the ratios at ages 50 and above tend to be similar. By the early 60s, the
gap between southern and northern Europe has shrinked. By now, the ratios between mortality quotients
tend to be within a factor of 2 across all ages, and even less at ages 50 and above.

:::

# Death rates evolution since WW II

::: {.callout-note}

### Question

Plot mortality quotients (column `q(x)`) for
both genders as a function of `Age` for  years `1946, 1956, ...` up to `2016` . Use aesthetics to distinguish years. You will need to categorize  the `Year`  column
(`forcats::` may be helpful).

1. Facet by `Gender`  and `Country`
2. Pay attention to axes labels, to legends. Assess logarithmic scales.

:::

::::: {.content-visible when-profile='solution'} 
 

```{r}
#| label: mx_1913_21_f

post_ww_II <- with(params, 
  seq(year_p, year_e, 10)
)

p <- life_tables |>
  filter(FALSE) |>
  ggplot() +
  aes(x=Age,
       y=`q(x)`,
       col=as_factor(Year),
       linetype=as_factor(Year)) +
  geom_smooth(se=FALSE, method="loess", span= .1, size=.2) +
  labs(colour="Year", linetype="Year")   +
  scale_y_log10() +
  facet_grid(rows=vars(Country), cols=vars(Gender))
```

```{r}
p_3 <- (p  +
  filter(life_tables,
    Year %in% post_ww_II, 
    Gender!="Both",
    Age < 90,
    Country %in% c('Spain', 'USA')) +
  labs(title="Mortality quotient per Age",
    subtitle = "Post WW II")
)
```

::: {.content-hidden unless-format="html"}

```{r}
p_3 |> 
  ggplotly()
```

:::

::: {.content-visible unless-format="html"}

```{r}
p_3 
```


```{r}
filter(life_tables,
    Year %in% post_ww_II, 
    Gender!="Both",
    Age < 11,
    Country %in% c('Spain', 'USA')) |>
ggplot() +
  aes(x=Age,
       y=`q(x)`,
       col=as_factor(Year),
       linetype=as_factor(Year)) +
  geom_point() +
#  geom_smooth(se=FALSE, method="loess", span= .1, size=.2) +
#  labs(colour="Year", linetype="Year")   +
  scale_y_log10() +
  facet_grid(rows=vars(Country), cols=vars(Gender))
```

:::

:::::


::: {.callout-note}

### Question

Write a function `ratio_mortality_rates` with signature
`function(df, reference_year=1946, target_years=seq(1946, 2016, 10))`
that takes as input:

- a dataframe with the same schema as `life_table`,
- a reference year `ref_year` and
- a sequence of years `target_years`

and that returns a dataframe with schema:


| Column Name | Column Type |
|:------------|:------------|
|  Year       | integer     |
|  Age        | integer     |
|  q(x)         | double      |
|  q(x).ref_year| double      |
|  Country    | factor      |
|  Gender     | factor      |

where `(Country, Year, Age, Gender)` serves as a _primary key_,
`q(x)` denotes the mortality quotient at `Age` for `Year` and `Gender` in `Country`
whereas `q(x)_ref_year` denotes mortality quotient at `Age` for argument `reference_year`
in `Country` for `Gender`.

:::

::::: {.content-visible when-profile='solution'} 
 

```{r}
#| label: ratio_mortality_rates
ratio_mortality_rates <- function(df,
                                  reference_year=1946,
                                  target_years=seq(1946, 2016, 10)){
  
  jbe <- join_by(Age, Gender, Country)
  
  right_df <- df |>
    filter(Year==reference_year) |>
    select(Age, Gender, Country, `q(x)`)
  
  df |> 
    filter(Year %in% target_years, Age <90) |>
    select(Age, Gender, Country, `q(x)`, Year) |>
    inner_join(right_df, by = jbe)
    
}
```

:::::

::: {.callout-note}

### Question

Draw plots displaying the ratio  $q_{x,t}/q_{x, 1946}$ for ages $x \in 1, \ldots, 90$ and year $t$ for $t \in 1946, \ldots, 2013$ where $q_{x,t}$ is the mortality quotient at age $x$ during year  $t$.

1. Handle both genders and all countries 
2. One properly facetted plot is enough.

:::

::::: {.content-visible when-profile='solution'} 
 
```{r}
#| label: com_ratio_mortality_rates
df_ratios <- life_tables |>
  filter(Gender !="Both") |>
  ratio_mortality_rates(reference_year=1948,
    target_years=seq(1948, 2013, 1))
```



```{r}
geom_smooth_line <- geom_smooth(method="loess",
  formula = y ~ x,
  se= FALSE,
  size =.2,
  span= .1
)
```

```{r}
#| label: full_comparison_template
#| warning: false
#| message: false
q <- df_ratios |>
  ggplot() +
  aes(x=Age,
      y=`q(x).x`/`q(x).y`) +
  geom_smooth_line +
  scale_y_log10() 

q1 <- q + 
  aes(linetype=as_factor(Year),
      col=as_factor(Year)) +
  ylab("Ratio of mortality rates, reference Year 1946") +
  labs(linetype="Year", col="Year") +
  scale_colour_brewer() +
  theme(legend.position = "none") +
  facet_grid(
    rows = vars(Country), 
    cols =vars(Gender),
    scales = "free_y"
  )

q1
```


```{r}
#| warning: false
#| message: false
#| 
q1 + (
  df_ratios |>
    filter(Age <= 20)
)

```


```{r}
#| label: frame_full_comparison
# 
qf <- q +
  aes(linetype=Country,
      frame=Year) +
  geom_smooth(method="loess",
              formula= y ~ x, 
              se= FALSE,
              size =.2,
              span= .1) +
  labs(linetype="Country") +
  facet_grid(rows=vars(Gender))
```

```{r}
qf |> ggplotly()
```

:::::

::: {.content-visible when-profile='solution'} 
 

```{r}
#| label: full_comparison
# 
with(params, 
     q + filter(df_ratios, 0 == (Year - year_p) %% 15) +
  ggtitle("Full comparison") +
  facet_grid(cols=vars(Gender), rows=vars(Country)))
```

:::


::: {.callout-note}

### Comment. 

During the last seventy years, death rates decreased at all ages in all seven countries.This progress has not been uniform across ages, genders and countries. Across most countries, infant mortality dramatically improved during the first post-war decade  while death rates at age 50 and above remained stable
until the mid seventies.

:::


# Trends

We noticed that mortality quotients did not evolve in the same way across all ages: first, the decay has been much more significant at low ages; second, the decay of mortality quotients at old ages (above 60) mostly took place  during the last four decades. It is worth digging separately at what happened for different parts of life.

::: {.callout}

### Question 

Plot mortality quotients at ages $0, 1, 5$  as a function of time. Facet by Gender and Country

:::

::::: {.content-visible when-profile='solution'} 
 
```{r}
#| label: infant_mort_rate
ages <- c(0, 1, 5)

p_children <- filter(life_tables, FALSE) |>
  ggplot(mapping=aes(x=Year, y=`q(x)`,
                     linetype=forcats::as_factor(Age),
                     shape=forcats::as_factor(Age),
                     col=forcats::as_factor(Age))) +
  geom_line(size=.2) +
  labs(linetype="Age", col="Age", shape="Age") +
  scale_y_log10() +
  scale_x_continuous(breaks=seq(1935,2020,5)) +
  facet_grid(cols=vars(Gender), rows=vars(Country))

p_children +
  filter(life_tables,
            Age %in% ages,
            Gender != "Both",
            Year %in% 1933:2013) +
  ggtitle("Infant and child, mortality rate",
            subtitle = "Hygiene, Vaccination, Antibiotics")
```
 
:::::


::: {.callout-note}

### Comment

All European countries achieved the same infant mortality rates after year 2000. The USA now lag behind.

During years 1940-1945, in the Netherlands and France, gains obtained before 1940 were reversed. Year 1945
was particularly difficult in the Netherlands.

:::

::: {.callout-note}

### Question

Plot mortality quotients at ages $15, 20, 40, 60$  as a function of time. Facet by `Gender` and `Country`

:::

::::: {.content-visible when-profile='solution'} 
 

```{r}
#| label: adult_mortality_rates
ages <- c(15, 20, 40, 60)

p_children +
  filter(life_tables,
         Age %in% ages,
         Gender != "Both",
         Year %in% 1935:2020) +
  ggtitle("Mortality rate at different ages")
```

:::::

::: {.callout-note}

### Comment. 

While death rates at ages 15 and 20 among women are close across all societies,  death
rates are higher at age 20 than at age 15 among men. In France, at age 20, death rates declined from 1945
until 1960, and then increased back to their initial level until 1980. Male death rates at age 60 started
to decline around 1980. Female death rates at age 60 declined steadily throughout the 7 decades.
Years 1940-1945 exhibit disruptions with different shapes and intensities in  Italy, France, England & Wales, and the Netherlands.

:::

::::: {.content-visible when-profile="solution"}

# Appendix 


```{r}
#| label: chunk-make-appendix
#| ref.label: !expr knitr::all_labels(appendix==TRUE)
#| echo: true
#| eval: false
#| code-fold: false
```


:::::